DeepPicar: A Low-cost Deep Neural Network-based Autonomous Car

### 2018-01-31 paper/background.tex
Changes to paper

The use of neural networks for end-to-end control of autonomous
car was first demonstrated in late 1980s~\cite{Pomerleau1989},
using a small 3-layer fully connected neural network; and subsequently
in a DARPA Autonomous Vehicle (DAVE) project in early
2000s~\cite{LeCun:04}, using a 6 layer convolutional neural network
(CNN); and most recently in NVIDIA's DAVE-2
project~\cite{Bojarski2016}, using a 9 layer CNN. In all of these projects,
the neural network models take raw image pixels as input and directly
produce steering control commands, bypassing all intermediary steps and
hand-written rules used in the conventional robotics control approach.  
NVIDIA's latest effort reports that their trained CNN
autonomously controls their modified cars on public roads without human
intervention~\cite{Bojarski2016}.

@@

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chips (SOCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with lots of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
For these reasons, finding a good computing platform is a
non-trivial task, one that requires a deep understanding of the
workloads and the hardware platform being utilized.

>>>>

The use of neural networks for end-to-end control of autonomous
car was first demonstrated in the late 1980s~\cite{Pomerleau1989},
using a small 3-layer fully connected neural network; and subsequently
in a DARPA Autonomous Vehicle (DAVE) project in early
2000s~\cite{LeCun:04}, using a 6 layer convolutional neural network
(CNN); and most recently in NVIDIA's DAVE-2
project~\cite{Bojarski2016}, using a 9 layer CNN. In all of these projects,
the neural network models take raw image pixels as input and directly
produce steering control commands, bypassing all intermediary steps and
hand-written rules used in the conventional robotics control approach.  
NVIDIA's latest effort reports that their trained CNN
autonomously controls their modified cars on public roads without human
intervention~\cite{Bojarski2016}.

@@

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chips (SoCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with lots of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
For these reasons, finding a good computing platform is a
non-trivial task, one that requires a deep understanding of the
workloads and the hardware platform being utilized.

### 2018-01-31 paper/comparison.tex
Changes to paper

In this section, we compare three computing platforms---the Raspberry
Pi 3, the Intel UP~\footnote{http://www.up-board.org/up/} and NVIDIA
Jetson
TX2~\footnote{http://www.nvidia.com/object/embedded-systems-dev-kits-modules.html}---from
the point of view of supporting end-to-end deep learning
based autonomous vehicles. 
Table~\ref{tbl:platforms} shows architectural features of the three
platforms~\footnote{The GPUs of the Raspberry Pi 3 and Intel
  Up are not used in evaluation due to the lack of software (TensorFlow)
support. Also, the two Denver cores in Tegra TX2 are not used in
evaluation due to TensorFlow issues.}.

@@

Figure~\ref{fig:sys_core} shows the average control loop completion
timing of the four system configurations we tested as a function of
the number of CPU cores used.
First, both the Intel UP and Jetson TX2 exhibit superior performance when
compared with the Raspberry Pi 3. 
When all four CPU cores are used, the Intel UP is 1.33X faster than
the Pi 3, while the TX2 (CPU) and TX2 (GPU) are 2.79X and 4.16X times faster,
respectively, than the Pi 3. 
As a result, they are all able to satisfy the 33.$\overline{\mbox{33}}$ ms 
WCET by a clear margin,
%%  (in the worst case, the UP Board would 
%% still meet it by $\sim$20 ms) --> where's the data showing this?
and, in case of TX2, 50 Hz or even 100 Hz real-time control is
feasible with the help of its GPU. Another observation is that TX2
(GPU) does not change much, as most of the neural network computation
is done at the GPU.

@@

The Intel UP board and Jetson TX2 also perform much better when multiple DNN models
are co-scheduled. Figure~\ref{fig:sys_model} shows the results of the
multi-model co-scheduling experiment. Once again, they can comfortably
satisfy 30 Hz real-time performance for all of the co-scheduled DNN control
loops, and in the case of the TX2 (GPU), 100 Hz real-time control is still
feasible. Given that the GPU must be shared among the co-scheduled DNN
models, the results suggest that the TX2's GPU has sufficient capacity to
accomodate multiple instances of the DNN model we tested.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observed dramatic execution time
increases, up to 11.6X, in Raspberry Pi 3 as we increased the number of
co-scheduled tasks. We also observe increased control loop execution
timing in the Intel Up and Jetson TX2, but the degree of the increase is 
not as dramatic as the Pi 3. Compared to their respective solo timings 
(i.e., the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.1X increases, respectively. This is somewhat suprising
because the Raspberry Pi 3's cores are in-order architecture based while
the cores in the Intel Up and NVIDIA TX2 are out-of-order architecture
based, and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3. However, 
the L2 miss rates of the platforms follow the same expected 
pattern, as can be seen in Figure~\ref{fig:sys_l2miss}. Namely, the 
percentage of L2 miss rates increases are more 
memory intensive co-runners are introduced. We found that the Pi 3 had 
slightly higher L2 miss rates than the UP and TX2.

>>>>

In this section, we compare three computing platforms---the Raspberry
Pi 3, the Intel UP~\footnote{http://www.up-board.org/up/} and NVIDIA
Jetson
TX2~\footnote{http://www.nvidia.com/object/embedded-systems-dev-kits-modules.html}---from
the point of view of supporting end-to-end deep learning
based autonomous vehicles. 
Table~\ref{tbl:platforms} shows architectural features of the three
platforms~\footnote{The GPUs of the Raspberry Pi 3 and Intel
  Up are not used in evaluation due to the lack of software (TensorFlow)
support. Also, the two Denver cores in the Tegra TX2 are not used in
evaluation due to TensorFlow issues.}.

@@

Figure~\ref{fig:sys_core} shows the average control loop completion
timing of the four system configurations we tested as a function of
the number of CPU cores used.
First, both the Intel UP and Jetson TX2 exhibit superior performance when
compared with the Raspberry Pi 3. 
When all four CPU cores are used, the Intel UP is 1.33X faster than
the Pi 3, while the TX2 (CPU) and TX2 (GPU) are 2.79X and 4.16X times faster,
respectively, than the Pi 3. 
As a result, they are all able to satisfy the 33.$\overline{\mbox{33}}$ ms 
WCET by a clear margin,
%%  (in the worst case, the UP Board would 
%% still meet it by $\sim$20 ms) --> where's the data showing this?
and, in the case of the TX2, 50 Hz or even 100 Hz real-time control is
feasible with the help of its GPU. Another observation is that the TX2
(GPU) performance does not change much, as most of the neural network 
computation is done at the GPU.

@@

The Intel UP board and Jetson TX2 also perform much better when multiple DNN models
are co-scheduled. Figure~\ref{fig:sys_model} shows the results of the
multi-model co-scheduling experiment. Once again, they can comfortably
satisfy 30 Hz real-time performance for all of the co-scheduled DNN control
loops, and in the case of the TX2 (GPU), 100 Hz real-time control is still
feasible. Given that the GPU must be shared among the co-scheduled DNN
models, the results suggest that the TX2's GPU has sufficient capacity to
accomodate multiple instances of the DNN models we tested.

@@

Finally, we compare the effect of co-scheduling memory bandwidth
intensive synthetic benchmarks on the DNN control loop timing. 
Figure~\ref{fig:sys_bench} shows the results.
As discussed in 
Section~\ref{sec:eval-memhog}, we observed dramatic execution time
increases, up to 11.6X, in Raspberry Pi 3 as we increased the number of
co-scheduled tasks. We also observe increased control loop execution
timing in the Intel Up and Jetson TX2, but the degree of the increase is 
not as dramatic as the Pi 3. Compared to their respective solo timings 
(i.e., the model runs on a single core in isolation), Intel UP suffers up to
2.3X execution time increase; TX2 (CPU) and TX2 (GPU) suffer up to
2.2X and 2.1X increases, respectively. This is somewhat suprising
because the Raspberry Pi 3's cores are in-order architecture based while
the cores in the Intel Up and NVIDIA TX2 are out-of-order architecture
based, and that the memory intensive tasks on out-of-order cores can
generate more memory traffic. We suspect that this is because the
memory subsystems in the Intel UP and TX2 platforms provide higher
bandwidth and fairness than the memory subsystem of the Pi 3. However, 
the L2 miss rates of the Pi 3 and TX2 platforms follow the same expected 
pattern, as can be seen in Figure~\ref{fig:sys_l2miss}. Namely, the 
percentage of L2 miss rates increases are more 
memory intensive co-runners are introduced. This was not the case for 
the UP board, where the model running on two cores generated the highest
percentage of L2 cache misses. This is most likely due to the CPU 
architecture of the UP's CPU which is Intel based, whereas both the Pi 3
and TX2 have ARM based CPUs.

### 2018-01-31 paper/evaluation.tex
Changes to paper

In this section, we experimentally analyze various real-time aspects
of the DeepPicar. This includes
(1) measurement based worst-case execution time (WCET) analysis of
deep learning inferencing,
(2) the effect of using multiple cores in accelerating inferencing,
(3) the effect of co-scheduling multiple deep neural network models,
and 
(4) the effect of co-scheduling memory bandwidth intensive co-runners.
(5) the effect of partitioning the shared L2 cache.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 46.30 ms on a single core to 22.86 ms on four
cores (over a 100\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 3 cores to 4 cores, the
improvement is mere 2.80 ms, or 12%. In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
TensorFlow's multicore implementation or if it's the model's inherent
characteristics.

@@

Finally, we run a single model alongside memory intensive co-runners 
again, but ensure that each task is given equal amounts of L2 cache space.
For this, we assign two colors to each task, regardless of the number of
tasks being scheduled. As a result, the entire cache is not utilized in
experiments with less than four tasks. The results can be seen in 
\ref{fig:palloc_bandwidth_exectime}. With zero or one co-runners, the DNN
performance doesnt't change relative to when the cache wasn't partitioned.
However, we find that with two or more co-runners, DNN performance becomes
worse. In the case of two co-runners, average control loop time increases
by 66.31 ms, or 26\%, and with three co-runners, the average time
is 265.64 ms, of 49\%, greater.
At the same time, though, the L2 miss rates of the model don't correlate 
to the timing increases seen, as is shown in 
\ref{fig:palloc_bandwidth_l2missrate}. Instead, we find that the number
of bus accesses increases when cache partitions was used, and they mirror
the timing increases of the model. These additional bus accesses are
most likely caused by the L2 prefetcher, as it can generate additional bus
accesses that aren't counted as L2 misses. As a result, we find that cache
partitioning is detrimental to the performance of the DNN so long as the 
L2 prefetcher is enabled (we have not yet found a way to disable it on
the Pi 3). Furthermore, no cache sensitivity was displayed by the model,
thus showing that the performance of the L2 cache is not vital for DNN 
performance when co-runners are introduced.

@@

Lastly, we find that partitioning the shared L2 cache provides 
practically no benefits to the real-time performance of the DNN. To the 
contrary, performance is negatively affected when fewer cores are used, 
due to less cache being available, and memory intensive co-runners are 
introduced, as they generate more memory contention. Furthermore, we 
find that performance remains consistent when multiple models are run 
simultaneously. As such, we conclude that the shared L2 cache is not 
essential to the DNN's real-time performance.

>>>>

In this section, we experimentally analyze various real-time aspects
of the DeepPicar. This includes
(1) measurement based worst-case execution time (WCET) analysis of
deep learning inferencing,
(2) the effect of using multiple cores in accelerating inferencing,
(3) the effect of co-scheduling multiple deep neural network models,
(4) the effect of co-scheduling memory bandwidth intensive co-runners,
and
(5) the effect of partitioning the shared L2 cache.

@@

Figure~\ref{fig:perf-vs-corecnt} shows the average execution time of
the control loop as we vary the number of cores used by
TensorFlow. As expected, as we assign more cores, the average execution
time decreases---from 46.30 ms on a single core to 22.86 ms on four
cores (over a 100\% improvement). However, the improvement is far from an ideal
linear scaling. In particular, from 3 cores to 4 cores, the
improvement is mere 2.80 ms, or 12\%. In short, we find that the
scalability of DeepPicar's deep neural network is not ideal on the
platform. We do not know whether it is due to the limitations of
TensorFlow's multicore implementation or if it's the model's inherent
characteristics.

@@

Finally, we run a single model alongside memory intensive co-runners 
again, but ensure that each task is given equal amounts of L2 cache space.
For this, we assign two colors to each task, regardless of the number of
tasks being scheduled. As a result, the entire cache is not utilized in
experiments with less than four tasks. The results can be seen in Figure
\ref{fig:palloc_bandwidth_exectime}. With zero or one co-runners, the DNN
performance doesnt't change relative to when the cache wasn't partitioned.
However, we find that with two or more co-runners, DNN performance becomes
worse. In the case of two co-runners, average control loop time increases
by 66.31 ms, or 26\%, and with three co-runners, the average time
is 265.64 ms, of 49\%, greater.
At the same time, though, the L2 miss rates of the model don't correlate 
to the timing increases seen, as is shown in 
\ref{fig:palloc_bandwidth_l2missrate}. Instead, we find that the number
of bus accesses increases when cache partitions are used, and they mirror
the timing increases of the model. This can be seen in Figure 
\ref{fig:palloc_bandwidth_bus}. These additional bus accesses are
most likely caused by the L2 prefetcher, as it can generate additional bus
accesses that aren't counted as L2 misses. As a result, we find that cache
partitioning is detrimental to the performance of the DNN so long as the 
L2 prefetcher is enabled (we have not yet found a way to disable it on
the Pi 3). Furthermore, no cache sensitivity was displayed by the model,
thus showing that the performance of the L2 cache is not vital for DNN 
performance when co-runners are introduced.

@@

Lastly, we find that partitioning the shared L2 cache provides 
practically no benefits to the real-time performance of the DNN. To the 
contrary, performance is constant regardless of the amount of cache 
space used. We also find that performance becomes worse when memory
intensive co-runners are introduced as they generate more memory 
contention when partitioning is enabled. Furthermore, we 
find that performance remains consistent when multiple models are run 
simultaneously. As such, we conclude that the shared L2 cache is not 
essential to the DNN's real-time performance.
